---
title: "Education Funding: FUNDEB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Education Funding: FUNDEB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette covers FUNDEB (Fundo de Manutencao e Desenvolvimento da Educacao
Basica), the main funding mechanism for basic education in Brazil. educabR
provides two functions to access FUNDEB data: resource distribution and
enrollment counts.

```{r setup}
library(educabR)
library(dplyr)
library(ggplot2)
```

## Overview

FUNDEB redistributes tax revenues to fund basic education across states and
municipalities. The data is split into two components:

| Function | Source | Description |
|----------|--------|-------------|
| `get_fundeb_distribution()` | STN (National Treasury) | Monthly transfers by state/municipality and funding source |
| `get_fundeb_enrollment()` | FNDE (OData API) | Enrollment counts used for funding calculation |

---

## FUNDEB Distribution

Resource distribution data shows how much money was transferred to each state
and municipality, broken down by funding source and month.

### Funding sources

FUNDEB aggregates revenue from multiple tax sources:

| Source | Description |
|--------|-------------|
| FPE | State Participation Fund |
| FPM | Municipal Participation Fund |
| IPI | Tax on Industrialized Products |
| ITR | Rural Property Tax |
| VAAF | Fiscal Value-Added (state portion) |
| VAAT | Total Value-Added |
| VAAR | FUNDEB Complementation |
| ICMS | State Sales Tax |
| IPVA | Vehicle Tax |
| ITCMD | Inheritance and Donation Tax |

### Downloading distribution data

```{r distribution-download}
# All distribution data for 2023
dist_2023 <- get_fundeb_distribution(year = 2023)

# Filter by state
dist_sp <- get_fundeb_distribution(year = 2023, uf = "SP")

# Filter by funding source
dist_fpm <- get_fundeb_distribution(year = 2023, source = "FPM")

# Filter by destination (states or municipalities)
dist_estados <- get_fundeb_distribution(
  year        = 2023,
  destination = "estados"
)
```

### Available years

Data is available from 2007 to 2026.

### Data structure

```{r distribution-structure}
dist <- get_fundeb_distribution(2023, uf = "SP")
glimpse(dist)
```

### Example analysis: Monthly transfers by source

```{r distribution-analysis}
dist <- get_fundeb_distribution(2023, uf = "SP")

# Total monthly transfers by funding source
monthly <- 
  dist |>
  group_by(fonte, mes) |>
  summarise(total = sum(valor, na.rm = TRUE), .groups = "drop")

ggplot(monthly, aes(x = mes, y = total / 1e6, fill = fonte)) +
  geom_col() +
  labs(
    title = "FUNDEB Transfers to Sao Paulo by Source (2023)",
    x     = "Month",
    y     = "Total (millions R$)",
    fill  = "Source"
    ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Example analysis: Comparing states

```{r distribution-states}
dist <- get_fundeb_distribution(2023)

# Total annual transfers by state
by_state <- 
  dist |>
  group_by(uf) |>
  summarise(total = sum(valor, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total)) |>
  head(10)

ggplot(by_state, aes(x = reorder(uf, total), y = total / 1e9)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  labs(
    title = "Top 10 States by FUNDEB Transfers (2023)",
    x     = NULL,
    y     = "Total (billions R$)"
    ) +
  theme_minimal()
```

---

## FUNDEB Enrollment

Enrollment data shows the number of students considered for FUNDEB funding
calculation, broken down by municipality, education type, school network,
and location.

### Downloading enrollment data

```{r enrollment-download}
# All enrollment data for 2023
mat_2023 <- get_fundeb_enrollment(year = 2023)

# Filter by state (applied at the API level for efficiency)
mat_sp <- get_fundeb_enrollment(year = 2023, uf = "SP")

# Limited rows for exploration
mat_sample <- get_fundeb_enrollment(year = 2023, n_max = 5000)
```

### Available years

Data is available from 2007 to 2026. Data is fetched from the FNDE OData API
with automatic pagination.

### Data structure

The enrollment data contains these columns:

| Column | Description |
|--------|-------------|
| `ano_censo` | Census year |
| `uf` | State code |
| `municipio` | Municipality name |
| `tipo_rede_educacao` | Education network type |
| `descricao_tipo_educacao` | Education type description |
| `descricao_tipo_ensino` | Teaching type description |
| `descricao_tipo_turma` | Class type description |
| `descricao_tipo_carga_horaria` | Class hours type |
| `descricao_tipo_localizacao` | Location type (urban/rural) |
| `qtd_matricula` | Number of enrollments |

```{r enrollment-structure}
mat <- get_fundeb_enrollment(2023, uf = "RJ")
glimpse(mat)
```

### Example analysis: Enrollment by education type

```{r enrollment-analysis}
mat <- get_fundeb_enrollment(2023, uf = "SP")

# Total enrollment by education type
by_type <- 
  mat |>
  group_by(descricao_tipo_educacao) |>
  summarise(total = sum(qtd_matricula, na.rm = TRUE), .groups = "drop") |>
  arrange(desc(total))

ggplot(by_type, aes(x = reorder(descricao_tipo_educacao, total), y = total / 1e3)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(
    title = "FUNDEB Enrollment by Education Type - SP (2023)",
    x     = NULL,
    y     = "Enrollments (thousands)"
    ) +
  theme_minimal()
```

### Example analysis: Urban vs rural enrollment

```{r enrollment-location}
mat <- get_fundeb_enrollment(2023)

# Compare urban vs rural
by_location <- 
  mat |>
  group_by(uf, descricao_tipo_localizacao) |>
  summarise(total = sum(qtd_matricula, na.rm = TRUE), .groups = "drop")

ggplot(by_location, aes(x = uf, y = total / 1e3, fill = descricao_tipo_localizacao)) +
  geom_col(position = "dodge") +
  labs(
    title = "FUNDEB Enrollment: Urban vs Rural by State (2023)",
    x     = "State",
    y     = "Enrollments (thousands)",
    fill  = "Location"
    ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Combining distribution and enrollment

You can combine both datasets to analyze per-student funding across states.

```{r combined}
# Total transfers by state
dist <- get_fundeb_distribution(2023)
transfers <- 
  dist |>
  group_by(uf) |>
  summarise(total_transfer = sum(valor, na.rm = TRUE), .groups = "drop")

# Total enrollment by state
mat <- get_fundeb_enrollment(2023)
enrollment <- 
  mat |>
  group_by(uf) |>
  summarise(total_students = sum(qtd_matricula, na.rm = TRUE), .groups = "drop")

# Per-student funding
funding <- 
  inner_join(transfers, enrollment, by = "uf") |>
  mutate(per_student = total_transfer / total_students) |>
  arrange(desc(per_student))

ggplot(funding, aes(x = reorder(uf, per_student), y = per_student)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "FUNDEB Per-Student Funding by State (2023)",
    x     = NULL,
    y     = "R$ per Student"
    ) +
  theme_minimal()
```
