---
title: "Getting Started with educabR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with educabR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The **educabR** package provides easy access to Brazilian public education data
from INEP, FNDE, and CAPES. With simple functions, you can download and process
data from 14 datasets:

**Basic Education:**  
- **IDEB** - Basic Education Development Index  
- **ENEM** - National High School Exam  
- **School Census** (Censo Escolar)  
- **SAEB** - Basic Education Assessment System  
- **ENCCEJA** - Youth and Adult Education Certification Exam  
- **ENEM by School** - ENEM results aggregated by school (2005-2015)  

**Higher Education:**  
- **Higher Education Census** (Censo da Educação Superior)  
- **ENADE** - National Student Performance Exam  
- **IDD** - Value-Added Indicator  
- **CPC** - Preliminary Course Concept  
- **IGC** - General Courses Index  

**Graduate Education:**  
- **CAPES** - Graduate programs, students, faculty, courses, and theses  

**Education Funding:**  
- **FUNDEB** - Resource distribution and enrollment data  

This vignette covers IDEB, ENEM, and the School Census. For other datasets, see:  
- `vignette("basic-education-assessments")` — SAEB, ENCCEJA, ENEM by School  
- `vignette("higher-education")` — Higher Education Census, ENADE, IDD, CPC, IGC, CAPES  
- `vignette("education-funding")` — FUNDEB  

## Installation
```{r install}
# Install from GitHub (development version)
# install.packages("remotes")
remotes::install_github("SidneyBissoli/educabR")
```

```{r setup}
library(educabR)
library(dplyr)
library(ggplot2)
```

## Configuring Cache

educabR caches downloaded files to avoid repeated downloads.
By default, it uses a temporary directory. To persist data across sessions:

```{r cache}
# Check current cache directory
get_cache_dir()

# Set a permanent directory
set_cache_dir("~/educabR_data")

# List cached files
list_cache()

# Clear cache (when needed)
clear_cache()
```

## IDEB - Basic Education Development Index

IDEB is the main indicator of basic education quality in Brazil,
combining test performance (SAEB) with grade promotion rates.

### Available years and levels

```{r ideb-available}
# See available combinations
list_ideb_available()
#> # A tibble: 24 x 3
#>     year level     stage
#>    <int> <chr>     <chr>
#>  1  2017 escola    anos_iniciais
#>  2  2017 escola    anos_finais
#>  3  2017 escola    ensino_medio
#>  4  2017 municipio anos_iniciais
#>  ...
```

### Downloading IDEB data

```{r ideb-download}
# IDEB by school - Early elementary (1st-5th grade)
ideb_schools <- get_ideb(
  year  = 2021,
  level = "escola",
  stage = "anos_iniciais"
)

# IDEB by municipality - High school
ideb_municipalities <- get_ideb(
  year  = 2023,
  level = "municipio",
  stage = "ensino_medio"
)

# Filter by state (faster)
ideb_sp <- get_ideb(
  year  = 2021,
  level = "escola",
  stage = "anos_iniciais",
  uf    = "SP"
)
```

### Data structure

```{r ideb-structure}
# View structure
glimpse(ideb_schools)
#> Rows: 63,529
#> Columns: 17
#> $ sg_uf                   <chr> "RO", "RO", "RO", ...
#> $ co_municipio            <dbl> 1100015, 1100015, ...
#> $ no_municipio            <chr> "Alta Floresta D'Oeste", ...
#> $ id_escola               <dbl> 11000023, 11000040, ...
#> $ no_escola               <chr> "EEEE ABNAEL MACHADO DE LIMA", ...
#> $ rede                    <chr> "Estadual", "Municipal", ...
#> $ vl_aprovacao_2021_si_4  <dbl> 93.3, 98.5, 100, ...
#> $ vl_indicador_rend_2021  <dbl> 0.92, 0.98, 1.00, ...
#> $ vl_nota_matematica_2021 <dbl> 5.2, 5.8, 6.1, ...
#> $ vl_nota_portugues_2021  <dbl> 5.4, 5.9, 6.0, ...
#> $ vl_nota_media_2021      <dbl> 5.3, 5.85, 6.05, ...
#> $ vl_observado_2021       <dbl> 4.9, 5.7, 6.1, ...
```

### Example analysis: Average IDEB by state

```{r ideb-analysis}
# Calculate average IDEB by state
ideb_by_state <- 
  ideb_schools |>
  filter(!is.na(vl_observado_2021)) |>
  group_by(sg_uf) |>
  summarise(
    n_schools   = n(),
    mean_ideb   = mean(vl_observado_2021, na.rm = TRUE),
    median_ideb = median(vl_observado_2021, na.rm = TRUE)
  ) |>
  arrange(desc(mean_ideb))

# Plot
ggplot(ideb_by_state, aes(x = reorder(sg_uf, mean_ideb), y = mean_ideb)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average IDEB by State - Early Elementary (2021)",
    x     = "State",
    y     = "Average IDEB"
    ) +
  theme_minimal()
```

### IDEB historical series

```{r ideb-series}
# Download historical series
ideb_history <- get_ideb_series(
  years = c(2017, 2019, 2021, 2023),
  level = "municipio",
  stage = "anos_iniciais"
)

# National trend
trend <- 
  ideb_history |>
  group_by(ano_ideb) |>
  summarise(mean_ideb = mean(vl_observado, na.rm = TRUE))

ggplot(trend, aes(x = ano_ideb, y = mean_ideb)) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(color = "darkgreen", size = 3) +
  labs(
    title = "IDEB Trend - Early Elementary",
    x     = "Year",
    y     = "National Average IDEB"
  ) +
  theme_minimal()
```

## ENEM - National High School Exam

ENEM is Brazil's largest exam, with millions of participants annually.
The microdata includes demographics, socioeconomic data, and scores.

### Downloading ENEM data

```{r enem-download}
# WARNING: Large files (1-3 GB)!
# Use n_max for initial exploration

# Sample for exploration
enem_sample <- get_enem(2023, n_max = 10000)

# Full data (takes a while!)
# enem_full <- get_enem(2023)
```

### Data structure

```{r enem-structure}
glimpse(enem_sample)
#> Rows: 10,000
#> Columns: 76
#> $ nu_inscricao      <dbl> 230001234567, ...
#> $ nu_ano            <dbl> 2023, 2023, ...
#> $ tp_faixa_etaria   <dbl> 3, 4, 2, ...
#> $ tp_sexo           <chr> "F", "M", "F", ...
#> $ tp_cor_raca       <dbl> 1, 3, 2, ...
#> $ nu_nota_cn        <dbl> 512.3, 489.1, ...
#> $ nu_nota_ch        <dbl> 598.2, 567.4, ...
#> $ nu_nota_lc        <dbl> 534.8, 502.1, ...
#> $ nu_nota_mt        <dbl> 478.9, 521.3, ...
#> $ nu_nota_redacao   <dbl> 720, 640, ...
```

### Performance analysis

```{r enem-analysis}
# Summary statistics for scores
enem_summary(enem_sample)

# Summary by sex
enem_summary(enem_sample, by = "tp_sexo")

# Average scores by race/ethnicity
scores_by_race <- 
  enem_sample |>
  filter(!is.na(nu_nota_mt)) |>
  mutate(
    race = case_when(
      tp_cor_raca == 0 ~ "Not declared",
      tp_cor_raca == 1 ~ "White",
      tp_cor_raca == 2 ~ "Black",
      tp_cor_raca == 3 ~ "Mixed race",
      tp_cor_raca == 4 ~ "Asian",
      tp_cor_raca == 5 ~ "Indigenous"
    )
  ) |>
  group_by(race) |>
  summarise(
    n          = n(),
    mean_math  = mean(nu_nota_mt, na.rm = TRUE),
    mean_essay = mean(nu_nota_redacao, na.rm = TRUE)
  )
```

### Item response data

```{r enem-itens}
# Data about exam questions
items <- get_enem_itens(2023)

glimpse(items)
#> $ co_item        <dbl> 1, 2, 3, ...
#> $ co_prova       <dbl> 1001, 1001, ...
#> $ tp_lingua      <dbl> 0, 0, ...
#> $ sg_area        <chr> "CN", "CN", ...
#> $ co_habilidade  <dbl> 1, 2, ...
```

## School Census (Censo Escolar)

The School Census is Brazil's main statistical survey on basic education,
covering all public and private schools.

### Downloading Census data

```{r censo-download}
# School data
schools_2023 <- get_censo_escolar(2023)

# Filter by state (faster)
schools_sp <- get_censo_escolar(2023, uf = "SP")

# Sample for exploration
schools_sample <- get_censo_escolar(2023, n_max = 1000)
```

### Data structure

```{r censo-structure}
# The census contains over 400 variables per school!
glimpse(schools_2023)
#> Rows: 217,625
#> Columns: 408
#> $ nu_ano_censo     <dbl> 2023, 2023, ...
#> $ sg_uf            <chr> "RO", "RO", ...
#> $ co_uf            <dbl> 11, 11, ...
#> $ no_municipio     <chr> "Porto Velho", ...
#> $ co_municipio     <dbl> 1100205, ...
#> $ no_entidade      <chr> "EEEE ABNAEL MACHADO DE LIMA", ...
#> $ co_entidade      <dbl> 11000023, ...
#> $ tp_dependencia   <dbl> 2, 3, 4, ...
#> $ tp_localizacao   <dbl> 1, 1, 1, ...
```

### Analysis: Schools by administrative type

```{r censo-analysis}
# Count by administrative type
schools_by_type <- 
  schools_2023 |>
  mutate(
    admin_type = case_when(
      tp_dependencia == 1 ~ "Federal",
      tp_dependencia == 2 ~ "State",
      tp_dependencia == 3 ~ "Municipal",
      tp_dependencia == 4 ~ "Private"
    )
  ) |>
  count(admin_type) |>
  mutate(pct = n / sum(n) * 100)

ggplot(schools_by_type, aes(x = reorder(admin_type, n), y = n, fill = admin_type)) +
  geom_col() +
  geom_text(aes(label = sprintf("%.1f%%", pct)), hjust = -0.1) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Number of Schools by Administrative Type (2023)",
    x = NULL,
    y = "Number of Schools"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Analysis: School infrastructure

```{r censo-infrastructure}
# Check infrastructure availability in public schools
infra <- 
  schools_2023 |>
  filter(tp_dependencia %in% c(2, 3)) |>  # Public schools only
  summarise(
    pct_internet      = mean(in_internet == 1, na.rm = TRUE) * 100,
    pct_library       = mean(in_biblioteca == 1, na.rm = TRUE) * 100,
    pct_computer_lab  = mean(in_laboratorio_informatica == 1, na.rm = TRUE) * 100,
    pct_sports_court  = mean(in_quadra_esportes == 1, na.rm = TRUE) * 100,
    pct_accessibility = mean(in_acessibilidade == 1, na.rm = TRUE) * 100
  )

print(infra)
#> # A tibble: 1 x 5
#>   pct_internet pct_library pct_computer_lab pct_sports_court pct_accessibility
#>          <dbl>       <dbl>            <dbl>            <dbl>             <dbl>
#> 1         78.3        42.1             35.2             48.7              32.1
```

## Best practices

### 1. Set a permanent cache
```{r best-practices-cache}
# At the start of each project
set_cache_dir("~/educabR_data")
```

### 2. Use samples for exploration
```{r best-practices-sample}
# Explore with a small sample first
test_data <- get_enem(2023, n_max = 1000)

# Then download the full data
# full_data <- get_enem(2023)
```

### 3. Filter by state when possible
```{r best-practices-filter}
# Faster than downloading everything and filtering afterwards
data_sp <- get_ideb(2021, level = "escola", stage = "anos_iniciais", uf = "SP")
```

### 4. Monitor memory usage
```{r best-practices-memory}
# Large files can consume a lot of RAM
# Use n_max or filter data after loading
data <- get_censo_escolar(2023)
data_filtered <- 
  data |>
  select(co_entidade, no_entidade, sg_uf, tp_dependencia)
rm(data)  # Free memory
gc()
```

## Data dictionary

For detailed information about variables, see the official documentation:

- **IDEB**: https://www.gov.br/inep/pt-br/areas-de-atuacao/pesquisas-estatisticas-e-indicadores/ideb
- **ENEM**: https://www.gov.br/inep/pt-br/acesso-a-informacao/dados-abertos/microdados/enem
- **School Census**: https://www.gov.br/inep/pt-br/acesso-a-informacao/dados-abertos/microdados/censo-escolar
- **SAEB**: https://www.gov.br/inep/pt-br/acesso-a-informacao/dados-abertos/microdados/saeb
- **ENADE**: https://www.gov.br/inep/pt-br/acesso-a-informacao/dados-abertos/microdados/enade
- **CAPES**: https://dadosabertos.capes.gov.br
- **FUNDEB/FNDE**: `https://www.fnde.gov.br`

## Known issues

1. **Slow downloads**: INEP servers can be slow. The package retries automatically up to 3 times.
2. **Large files**: ENEM and School Census files can be several GB. Use `n_max` for testing.
3. **Encoding**: Some files have encoding issues. The package handles this automatically.

## Contributing

Found a bug or want to suggest improvements? Open an issue on GitHub:
https://github.com/SidneyBissoli/educabR/issues
