---
title: "Basic Education Assessments: SAEB, ENCCEJA, and ENEM by School"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Education Assessments: SAEB, ENCCEJA, and ENEM by School}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

This vignette covers three basic education assessment datasets available in
educabR. For IDEB, ENEM, and the School Census, see `vignette("getting-started")`.

```{r setup}
library(educabR)
library(dplyr)
library(ggplot2)
```

## SAEB - Basic Education Assessment System

SAEB (Sistema de Avaliacao da Educacao Basica) is a biennial assessment that
measures student performance in Portuguese and Mathematics across Brazilian
basic education. It is one of the components used to calculate IDEB.

### Available data types

SAEB microdata includes four perspectives:

| Type | Description |
|------|-------------|
| `"aluno"` | Student-level results (scores, responses) |
| `"escola"` | School questionnaire data |
| `"diretor"` | Principal questionnaire data |
| `"professor"` | Teacher questionnaire data |

### Downloading SAEB data

```{r saeb-download}
# Student performance data
saeb_students <- get_saeb(year = 2023, type = "aluno")

# School questionnaire
saeb_schools <- get_saeb(year = 2023, type = "escola")

# Use n_max for exploration
saeb_sample <- get_saeb(year = 2023, type = "aluno", n_max = 5000)
```

### Available years

SAEB is conducted every two years: 2011, 2013, 2015, 2017, 2019, 2021, 2023.

```{r saeb-years}
# 2021 data is split by education level
saeb_fund <- get_saeb(
  year  = 2021,
  type  = "aluno",
  level = "fundamental_medio"
)

saeb_infantil <- get_saeb(
  year  = 2021,
  type  = "aluno",
  level = "educacao_infantil"
)
```

### Example analysis: Score distribution

```{r saeb-analysis}
# Explore student scores
saeb_sample <- get_saeb(2023, type = "aluno", n_max = 10000)

# Score distribution by subject
saeb_sample |>
  filter(!is.na(proficiencia_mt)) |>
  ggplot(aes(x = proficiencia_mt)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "SAEB 2023 - Mathematics Proficiency Distribution",
    x     = "Mathematics Score",
    y     = "Count"
    ) +
  theme_minimal()
```

---

## ENCCEJA - Youth and Adult Education Certification

ENCCEJA (Exame Nacional para Certificacao de Competencias de Jovens e Adultos)
provides certification for elementary and high school equivalency. It covers
four knowledge areas: Natural Sciences, Mathematics, Portuguese, and Social
Sciences.

### Downloading ENCCEJA data

```{r encceja-download}
# Download ENCCEJA microdata
encceja_2023 <- get_encceja(year = 2023)

# Sample for exploration
encceja_sample <- get_encceja(year = 2023, n_max = 5000)
```

### Available years

ENCCEJA data is available from 2014 to 2024.

```{r encceja-structure}
# Explore the data structure
glimpse(encceja_sample)
```

### Example analysis: Participation by state

```{r encceja-analysis}
encceja_2023 <- get_encceja(2023, n_max = 50000)

# Count participants by state
participants_by_state <- 
  encceja_2023 |>
  count(sg_uf_prova, sort = TRUE) |>
  head(10)

ggplot(participants_by_state, aes(
  x = reorder(sg_uf_prova, n),
  y = n
)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "ENCCEJA 2023 - Top 10 States by Participation",
    x     = "State",
    y     = "Number of Participants"
    ) +
  theme_minimal()
```

---

## ENEM by School (2005-2015)

ENEM by School (ENEM por Escola) provides ENEM results aggregated at the school
level. This dataset covers 2005 to 2015 in a single bundled file and was
**discontinued after 2015**.

### Downloading the data

Unlike other datasets, this function has no `year` parameter — it downloads
the entire 2005-2015 dataset at once.

```{r enem-escola-download}
# Download all ENEM by School data (2005-2015)
enem_escola <- get_enem_escola()

# Sample for exploration
enem_escola_sample <- get_enem_escola(n_max = 5000)
```

### Data structure

```{r enem-escola-structure}
glimpse(enem_escola_sample)
```

### Example analysis: School performance trends

```{r enem-escola-analysis}
enem_escola <- get_enem_escola()

# Average scores over time (public vs private)
trend <- 
  enem_escola |>
  filter(!is.na(nu_media_tot)) |>
  group_by(nu_ano, tp_dependencia_adm_escola) |>
  summarise(
    mean_score = mean(nu_media_tot, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  mutate(
    admin_type = case_when(
      tp_dependencia_adm_escola == 1 ~ "Federal",
      tp_dependencia_adm_escola == 2 ~ "State",
      tp_dependencia_adm_escola == 3 ~ "Municipal",
      tp_dependencia_adm_escola == 4 ~ "Private"
    )
  )

ggplot(trend, aes(x = nu_ano, y = mean_score, color = admin_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "ENEM Average Score by School Type (2005-2015)",
    x     = "Year",
    y     = "Average Total Score",
    color = "School Type"
    ) +
  theme_minimal()
```
